---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# lsr

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/lsr)](https://CRAN.R-project.org/package=lsr)
[![Codecov test coverage](https://codecov.io/gh/djnavarro/lsr/graph/badge.svg)](https://app.codecov.io/gh/djnavarro/lsr)
<!-- badges: end -->

**lsr** is a companion package to the textbook [*Learning Statistics with R*](https://learningstatisticswithr.com). It provides beginner-friendly wrappers for common statistical procedures — t-tests, chi-square tests, effect sizes, correlation matrices, and basic data manipulation — with output designed to be readable by students encountering statistics for the first time.

## Example

Here is an independent-samples t-test comparing extra sleep between two drug groups in the built-in `sleep` dataset:

```{r example}
library(lsr)
independentSamplesTTest(formula = extra ~ group, data = sleep)
```

Compared to base R's `t.test()`, the output labels every component in plain English and automatically reports Cohen's *d* alongside the test result.

## Where to go next

- [**Guided overview**](https://lsr.djnavarro.net/articles/overview.html) — a hands-on introduction for students and beginners, working through descriptive statistics, reshaping data, and hypothesis testing with a single example dataset.
- [**Critical commentary**](https://lsr.djnavarro.net/articles/commentary.html) — an honest account of the package's limitations and pointers to better tools for users who have outgrown it.
- [**Reference**](https://lsr.djnavarro.net/reference/) — documentation for all 29 functions, organised by topic.

## Installation

Install the released version from [CRAN](https://CRAN.R-project.org):

```r
install.packages("lsr")
```

Or the development version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("djnavarro/lsr")
```
