---
title: "A guided overview of lsr"
description: >
  A hands-on introduction to the lsr package for students and readers of
  Learning Statistics with R.
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Why this package exists

Learning statistics and learning R at the same time is genuinely hard. One of
the things that makes it harder than it needs to be is that base R's output
was designed for researchers who already understand what they are looking for.
When you run `t.test()` for the first time, the output tells you everything
you need, but it takes some experience to know which number is which.

The **lsr** package is a companion to the textbook
[*Learning Statistics with R*](https://learningstatisticswithr.com) (Navarro,
2015). It wraps the most common statistical procedures in functions that
produce more readable output — labels in plain English, effect sizes included
by default, and hypotheses stated explicitly so you can check that you ran
the right test. The goal is to let you focus on understanding the statistics
rather than decoding the output.

This article walks you through the main functions using a single dataset. By
the time you reach the end you will have seen everything the package has to
offer.

## Getting set up

Load the package with `library()`:

```{r setup}
library(lsr)
```

### The example dataset

Throughout this article we will use data from an imaginary study. Thirty
students from three tutorial groups (A, B, and C) sat an end-of-term exam.
Before the exam they attended a one-hour study-skills workshop; we have their
score on a short practice test both before (`score_pre`) and after
(`score_post`) the workshop. The dataset also records how many hours per week
each student typically studies and whether they passed the exam (a score of 65
or above counts as a pass).

```{r create-data}
set.seed(250)

n <- 30
students <- data.frame(
  id          = 1:n,
  group       = factor(rep(c("A", "B", "C"), each = 10)),
  gender      = factor(sample(c("female", "male"), n, replace = TRUE)),
  score       = pmax(40, pmin(100, c(
    round(rnorm(10, mean = 62, sd = 7)),  # group A
    round(rnorm(10, mean = 68, sd = 7)),  # group B
    round(rnorm(10, mean = 74, sd = 7))   # group C
  ))),
  study_hours = NA_real_,
  score_pre   = NA_integer_,
  score_post  = NA_integer_
)
students$study_hours <- round(students$score / 10 + rnorm(n, 0, 0.8), 1)
students$score_pre   <- pmax(30L, round(students$score - rnorm(n, mean = 8, sd = 3)))
students$score_post  <- students$score
students$passed      <- factor(ifelse(students$score >= 65, "yes", "no"))
```

Here are the first few rows:

```{r preview-data}
head(students)
```

### Inspecting the workspace with `who()`

Once you have some objects in memory it can be useful to get a quick overview
of what you are working with. `who()` lists every object in your workspace
along with its type and size:

```{r who}
who()
```

Each row shows the name, class, number of rows and columns (for data frames),
and the size in memory. This is especially handy early in a session when you
are not yet sure what is loaded.

---

## Descriptive statistics

### Confidence interval for the mean: `ciMean()`

`ciMean()` computes a confidence interval for the population mean. Here we
estimate the interval for exam scores:

```{r ciMean}
ciMean(students$score)
```

The output is a 95% interval by default. You can change this with the
`conf.level` argument — for example, `ciMean(students$score, conf.level = 0.99)`
for a 99% interval.

### Average absolute deviation: `aad()`

The average absolute deviation is a robust measure of spread: how far do
values typically sit from the mean?

```{r aad}
aad(students$score)
```

### Most common value: `modeOf()`

`modeOf()` returns the most frequently occurring value in a vector. For
categorical variables this is the modal category:

```{r modeOf}
modeOf(students$passed)
```

More students passed than failed (19 out of 30), so "yes" is the mode.

### Correlation matrix: `correlate()`

`correlate()` produces a correlation matrix for the numeric columns of a data
frame. You can also pass a single data frame and it will select the numeric
columns automatically, or supply a subset of columns:

```{r correlate}
correlate(students[, c("score", "study_hours")])
```

Students who study more tend to score higher — a correlation of 0.64. The
diagonal is left blank because a variable's correlation with itself is always
exactly 1, which carries no information.

---

## Reshaping data

Many statistical functions in R expect data in *long* format (one observation
per row), but data are sometimes collected or stored in *wide* format (one
participant per row, with repeated measurements in separate columns). The
`wideToLong()` and `longToWide()` functions convert between the two.

Our `students` data frame has two score columns — `score_pre` and
`score_post` — which follow the naming convention `measurement_condition`.
`wideToLong()` uses that underscore separator to split the column name into
a measurement variable (`score`) and a within-subject condition variable
(`time`):

```{r wideToLong}
scores_wide <- students[, c("id", "group", "score_pre", "score_post")]
scores_long <- wideToLong(data = scores_wide, within = "time", sep = "_")
head(scores_long)
```

The long format has one row per student per time point instead of one row per
student. To go the other way, `longToWide()` takes the long-format data and a
formula telling it which column to spread across conditions:

```{r longToWide}
scores_back <- longToWide(data = scores_long, formula = score ~ time, sep = "_")
head(scores_back)
```

The two data frames contain the same information, just arranged differently.

---

## Hypothesis tests

lsr provides wrapper functions for six common hypothesis tests. Each one
prints a structured summary that states the hypotheses, gives descriptive
statistics, and reports the test result alongside an effect size.

### One-sample *t*-test: `oneSampleTTest()`

A one-sample *t*-test asks whether the population mean is equal to some
specified value. Here we test whether the true mean exam score differs from 65
(the pass mark):

```{r oneSampleTTest}
oneSampleTTest(x = students$score, mu = 65)
```

The mean score (67.8) is above 65, but the *p*-value of 0.060 is above the
conventional 0.05 threshold. This is a borderline result — the evidence
suggests scores may be above the pass mark on average, but the sample of 30
students is small enough that we cannot be confident. Notice that the output
includes Cohen's *d* (0.36), which tells us the effect size is small to
moderate.

### Independent-samples *t*-test: `independentSamplesTTest()`

An independent-samples *t*-test compares means between two separate groups.
We compare groups A and B to see if they have different average scores:

```{r independentSamplesTTest}
ab <- droplevels(students[students$group %in% c("A", "B"), ])
independentSamplesTTest(formula = score ~ group, data = ab)
```

The mean score in group B (68.0) is about 6.9 points higher than group A
(61.1), and the difference is clearly significant (*p* = 0.002). Cohen's *d*
of 1.62 indicates a large effect — the groups differ by considerably more than
one pooled standard deviation.

### Paired-samples *t*-test: `pairedSamplesTTest()`

A paired *t*-test compares two measurements from the same participants. We
use the long-format `scores_long` data created earlier to test whether scores
improved after the study workshop:

```{r pairedSamplesTTest}
scores_long$id <- factor(scores_long$id)
pairedSamplesTTest(formula = score ~ time, data = scores_long, id = "id")
```

Scores improved by an average of 8.2 points after the workshop, and the
difference is highly significant (*p* < 0.001). The large Cohen's *d* of 3.19
reflects that the improvement is very consistent across students — the
standard deviation of the difference (about 2.6 points) is much smaller than
the mean change.

### Chi-square test of association: `associationTest()`

`associationTest()` tests whether two categorical variables are independent of
one another. Here we ask whether passing rates differ between female and male
students:

```{r associationTest}
associationTest(formula = ~ gender + passed, data = students)
```

The output shows both the observed and expected contingency tables, making it
easy to see where the data diverge from the null hypothesis. Male students
passed at a higher rate (81%) than female students (43%), but with only 30
students and a *p*-value of 0.072, we do not have strong enough evidence to
conclude there is a real association. This is a good example of a result where
the pattern looks interesting but the sample is too small to be sure.

### Goodness-of-fit test: `goodnessOfFitTest()`

A goodness-of-fit test asks whether the observed frequencies of a categorical
variable match a set of expected proportions. Here we check whether students
are distributed equally across the three tutorial groups:

```{r goodnessOfFitTest}
goodnessOfFitTest(x = students$group)
```

By design, the three groups are perfectly balanced (10 students each), so
*p* = 1. In a real study you would use this test when you have a theoretical
expectation about the distribution — for example, testing whether the
proportion of students achieving each grade matches a historical benchmark.

### Post-hoc pairwise *t*-tests: `posthocPairwiseT()`

When an ANOVA finds an overall difference across three or more groups, you
often want to know *which* pairs of groups differ. `posthocPairwiseT()` runs
all pairwise *t*-tests with a correction for multiple comparisons. It takes
the result of `aov()` as its input:

```{r posthocPairwiseT}
score_model <- aov(score ~ group, data = students)
posthocPairwiseT(score_model)
```

All three pairs of groups differ significantly after the Holm correction. The
smallest *p*-value is for the A versus C comparison, which is unsurprising
given that group A has a mean of 61 and group C has a mean of 74.

---

## Effect sizes

Effect sizes tell you *how large* a difference is, not just whether it is
statistically significant. lsr provides three effect size functions.

### Cohen's *d*: `cohensD()`

Cohen's *d* measures the size of a difference between two groups in standard
deviation units. We already saw it appear automatically in the *t*-test
output, but you can also compute it directly:

```{r cohensD}
cohensD(formula = score ~ group, data = ab)
```

A value of 1.62 is a large effect by conventional benchmarks (small ≈ 0.2,
medium ≈ 0.5, large ≈ 0.8).

### Eta-squared: `etaSquared()`

Eta-squared measures the proportion of variance in the outcome explained by a
categorical predictor. It is the ANOVA equivalent of R-squared in regression.
Pass the `aov()` result:

```{r etaSquared}
etaSquared(score_model)
```

Eta-squared of 0.495 means the tutorial group explains about 49% of the
variance in exam scores. That is a large effect — the group a student belongs
to is strongly predictive of their score.

### Cramér's *V*: `cramersV()`

Cramér's *V* is an effect size for the chi-square test of association. It
runs from 0 (no association) to 1 (perfect association). Pass a contingency
table:

```{r cramersV}
cramersV(table(students$gender, students$passed))
```

A value of 0.33 suggests a moderate association between gender and passing
status, consistent with the pattern we saw in the contingency table earlier —
even though the chi-square test itself did not reach significance with this
small sample.

---

## Where to go next

You have now seen all the major functions in lsr in action. A few places to
continue from here:

- **The textbook.** [*Learning Statistics with R*](https://learningstatisticswithr.com)
  explains the theory behind every procedure covered in this article, and uses
  lsr throughout. It is freely available online.
- **The reference pages.** The [Reference](https://lsr.djnavarro.net/reference/)
  section of the documentation site has a full description of every function
  and its arguments.
- **The critical commentary.** If you are an intermediate R user — or a student
  who has finished an introductory course and wants to know what comes next —
  the [commentary article](https://lsr.djnavarro.net/articles/commentary.html)
  gives an honest assessment of lsr's limitations and points to more capable
  alternatives.
